% $Id$
% Author:	Daniel Bosk <daniel.bosk@miun.se>
% Date:		17 Dec 2012
\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[swedish,english]{babel} % english default language
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{prettyref,varioref}
\usepackage[numbers,square]{natbib}
\usepackage{color}
\usepackage{listings}
\bibliographystyle{plainnat}

\title{Laboratory Report on Paging Algorithms}
\author{Daniel Bosk}
\date{\today}

%\lstset{style=term}

\begin{document}
\maketitle
\begin{abstract}
	An abstract summarizing the report in at most 200 words.
\end{abstract}


\section{Introduction}
\label{sec:Intro}
A gentle introduction for the reader \dots


\section{Theory}
\label{sec:Theory}
A short descriptions of the algorithms in question \dots with references to
\citet{Silberschatz2009osc,Silberschatz2013intl,Silberschatz2013osc}.


\section{Results}
\label{sec:Results}
The hard-earned results \dots

Reference string X generated Y page-faults of Z possible, see the data in 
\prettyref{app:Data}.


\section{Analysis}
\label{sec:Analysis}
Analyze and try to explain the results.
Draw conclusions which are then summarized in \prettyref{sec:Conclusion}.


\section{Conclusion}
\label{sec:Conclusion}
Here you summarize the following:
\begin{enumerate}
	\item Answer whether the algorithms are correctly implemented or not?
	\item Provide a reference string which generates a minimum page-fault rate 
		for each algorithm.
	\item Provide a reference string which generates a maximum page-fault rate 
		for each algorithm.
	\item Provide evidence of Belady's anomaly for the algorithms suffering from 
		it.
\end{enumerate}


\bibliography{literature}

\appendix

\section{Data}
\label{app:Data}
Reference strings and corresponding output are given here.
The first FIFO-experiment yielded the following data.
The reference string:
\lstinputlisting{fifo-refstr.txt}
And the resulting output from the program:
\lstinputlisting{fifo-result.txt}

\end{document}
